Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2024-12-05 paper/paper.tex
paper

\begin{abstract}
  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
  innermost loops. Other areas, such as outer loop vectorization, are also in development.
  In this paper we give an introduction into VPlan and summarize its latest development results.
\end{abstract}

@@

This system, however, had quite a few limitations, as the loop vectorizer could only handle
innermost loops and neither outer loops, complex control flow, nor non-inlined functions. 
Additionally, while multiple different vectorizations for the same scalar code 
would be possible, the current vectorizers working directly on the LLVM IR could not model
and compare the costs of such different vectorization approaches.

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
control flow.

The auto-vectorization pass would create multiple different VPlans, based on initial legality 
checks. Each VPlan represents a different vectorization approaches, by modeling the
underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
transformations can then optimize these abstracted control flows based on a cost model.
After that, the best VPlan, and with it the best vectorization approach, is chosen by the cost model.
This VPlan is then executed, by going over its own vectorized control flow representation and
materializing it back into the underlying LLVM IR.

@@

The rest of this paper is organized as follows. We first explain different vectorization
strategies and vectorization constraints in general and describe the original auto-vectorization
approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
its most important individual components and gives an outlook on future developments. In 
Section~\ref{sec:relatedwork} we highlight related work to LLVM's auto-vectorization. Concluding,
we summarize VPlan, its current state and potential future.

@@

\paragraph{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control flow but operating on multiple values at once through vector registers.

@@

Uncountable loops can typically also not be vectorized, due to their unknown number of iterations.

Finally some instructions like integer division do inherently not have a vectorized instruction equivalent.

@@

Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
and permutation instructions. Control flow with many non-adjacent memory loads/stores or 
operations that combine different vector elements, such as reductions, can decrease vectorization 
performance.

@@

\subsection{Auto-Vectorization in LLVM}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize~\cite{llvmvec,llvmhistorystate}. Auto-Vectorization is part of LLVM's 
middle-end and runs after module simplification has taken place.
In general, these vectorizers take in scalar LLVM IR and transform it into optimized vectorized 
LLVM IR.

@@

The goal of Intels initial VPlan proposal was to unite any 
auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a single flexible
system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR by using its own 
hierarchical control-flow-graph (H-CFG). This allows for iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan would be able to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

This planned systems is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
Phase would check if the scalar code would be legal to vectorize at all, and if so, 
initial VPlans would be constructed. In the following Planning Phase, these initial 
VPlans would be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
The VPlan's cost model would supervise those transformations. Finally, the best VPlan 
would be chosen and executed, modifying the underlying IR.

@@

\subsection{VPlan Structure}
Before explaining how far this system is already implemented in LLVM we will go over
the different elements of a \texttt{VPlan} in more detail.

The purpose of a single VPlan object is to model a candidate for vectorization. 
Such a single candidate can potentially represent multiple different vectorized loops in the end. 
For example, one VPlan might model a general vectorization approach for multiple different 
Vectorization Factors (VFs) and multiple different Unroll Factors (UFs) if applicable.

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main element of a single VPlan.
The nodes of this H-CFG are one of three types:

\begin{itemize}
  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes}, similar to 
  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
  instructions within the final vectorized loop body.
  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
  IR \texttt{BasicBlock}. These nodes are useful for modeling the scalar parts around the 
  final vectorized loops such as the loop preheader and the epilogue loop.
\end{itemize}

@@

The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
infront of them indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related. The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). This block 
is used to decide if the scalar epilogue loop is to be executed following the vectorized loop.

@@

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen quite some development, any 
outer loop vectorization features are still experimental and need to be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

\subsubsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a number of extra conditions to check.

Some of the most important legality checks done are:
\begin{itemize}
  \item The scalar loop's control flow must not contain indirect branches or multiple backedges.
  An indirect branch is a branch that jumps to a value evaluated at runtime. A backedge in a loop 
  is an edge from the loop back to it. 
  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
  statement must not exit the loop, all instructions of such a block can be executed speculatively 
  (e.g. via masking)
  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
  variable, or a fixed-order recurrence.
  \item The loop must not contain function calls outside of standard math library calls.
  \item Types must be vectorizable. Vectorizable types are integers, floating point types, and 
  pointers.
  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
\end{itemize}

One can find a lot more details about these legality checks and their implementations by looking 
at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.

\subsubsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by first creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After finishing modeling each scalar IR instruction into the VPlan, a number of VPlan transformations 
are executed, optimizing the initialized VPlan.

After the initial VPlans have been optimized the best Vectorization Factor is computed. This is done 
by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
this step computes a optimal VF of $1$, it means that any form of vectorization is not profitable.

\paragraph{Cost Model}
Currently LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instructions are then generally just lookup tables influenced by the underlying 
compile target.

\subsubsection{Phase 3: Execution}
After the best VPlan has been chosen it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan some final transformations are made that depend on a specifically chosen VF and UF.
First a vectorized code skeleton is created in the IR, then the VPlan is executed. Upon execution 
the H-CFG of the VPlan is walked and each node is executed, which in turn executes all of its 
\texttt{VPRecipes} in order.

Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and for branches 
the general branch condition per block.

@@

In case that the epilogue loop could also be further vectorized, it will be in a second 
pass by creating VPlans for it, finding the best one and executing it.

@@

For example, function vectorization can be transformed into loop vectorization by creating a loop 
around the function's body and vectorizing it and vice versa. 
The same applies for SLP Vectorization and loop vectorization by unrolling a loop.

RV's goal is to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
A region is defined as any control flow that has a single entry and one or more non-divergent exits.
Non-divergent exits mean that starting from the same entry, control flow must not diverge to different
exits.
Such a region can then be for example an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.

To achieve this, RV implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control-flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

@@

\section{Summary and Future Work}
\label{sec:summary}
In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model vectorization 
specific instructions. This abstraction allows for efficiently exploring many different optimization 
decisions at once, keeping the underlying implementation flexible and modular. In the end a cost model, 
currently still working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
and choose the most optimal one. Finally this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
back into LLVM IR.

>>>>

\begin{abstract}
  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
  innermost loops. Other areas, such as outer loop vectorization, are also in development.
  In this paper, we introduce VPlan and summarize its latest development results.
\end{abstract}

@@

This system, however, had quite a few limitations, as the loop vectorizer could only handle
innermost loops and neither outer loops, complex control-flow, nor non-inlined functions. 
Additionally, while multiple different vectorizations for the same scalar code 
would be possible, the current vectorizers working directly on the LLVM IR could not model
and compare the costs of such different vectorization approaches.

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
control-flow.

The auto-vectorization pass would create multiple different VPlans based on initial legality 
checks. Each VPlan represents a different vectorization approach, by modeling the
underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
transformations can then optimize these abstracted control-flows based on a cost model.
After that, the best VPlan, and with it, the best vectorization approach is chosen by the cost model.
This VPlan is then executed by going over its own vectorized control flow representation and
materializing it back into the underlying LLVM IR.

@@

The rest of this paper is organized as follows. First, we explain different vectorization
strategies and constraints in general and describe the original auto-vectorization
approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
its most important individual components, and gives an outlook on future developments. In 
Section~\ref{sec:relatedwork}, we highlight work related to LLVM's auto-vectorization. Concluding,
we summarize VPlan, its current state, and its potential future.

@@

\paragraph{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control-flow but operating on multiple values at once through vector registers.

@@

Uncountable loops can typically also not be vectorized due to their unknown number of iterations.

Finally, some instructions, such as integer division, do inherently not have a vectorized instruction equivalent.

@@

Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
and permutation instructions. control flow with many non-adjacent memory loads/stores or 
operations that combine different vector elements, such as reductions, can decrease vectorization 
performance.

@@

\subsection{Auto-Vectorization in LLVM}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize~\cite{llvmvec,llvmhistorystate}. Auto-Vectorization is part of LLVM's 
middle-end and runs after module simplification.
These vectorizers take scalar LLVM IR and transform it into optimized vectorized 
LLVM IR.

@@

Intel's initial VPlan proposal aimed to unite any 
auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a 
single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

This planned system is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
Phase would check if the scalar code would be legal to vectorize at all, and if so, 
initial VPlans would be constructed. In the following Planning Phase, these initial 
VPlans would be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
The VPlan's cost model would supervise those transformations. Finally, the best VPlan 
would be chosen and executed, modifying the underlying IR.

@@

\subsection{VPlan Structure}
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

The purpose of a single VPlan object is to model a candidate for vectorization. 
Such a single candidate can potentially represent multiple different vectorized loops. 
For example, one VPlan might model a general vectorization approach for various 
Vectorization Factors (VFs) and Unroll Factors (UFs).

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan.
The nodes of this H-CFG are one of three types:

\begin{itemize}
  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes} similar to 
  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
  instructions within the final vectorized loop body.
  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
  IR \texttt{BasicBlock}. These nodes help model the scalar parts around the 
  final vectorized loops, such as the loop preheader and the epilogue loop.
\end{itemize}

@@

The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
in front of them indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related. The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
This block decides if the scalar epilogue loop will be executed following the vectorized loop.

@@

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

\subsubsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

Some of the most important legality checks done are:
\begin{itemize}
  \item The scalar loop's control flow must not contain indirect branches or multiple back-edges.
  An indirect branch is a branch that jumps to a value evaluated at runtime. A back-edge in a loop 
  is an edge from the loop back to it. 
  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
  statement must not exit the loop. All instructions of such a block can be executed speculatively 
  (e.g. via masking)
  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
  variable, or a fixed-order recurrence.
  \item The loop must not contain function calls outside of standard math library calls.
  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
\end{itemize}

One can find more details about these legality checks and their implementations by looking 
at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.

\subsubsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

After the initial VPlans have been optimized, the best Vectorization Factor is computed. This is done 
by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
this step computes an optimal VF of $1$, any vectorization is not profitable compared to the 
scalar version.

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
compile target.

\subsubsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which executes all 
of its \texttt{VPRecipes} in order.

Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and, for branches, 
the general branch condition per block.

@@

If the epilogue loop could be further vectorized, it would be in a second pass by creating VPlans, 
finding the best one, and executing it.

@@

For example, function vectorization can be transformed into loop vectorization by creating a loop 
around the function's body and vectorizing it and vice versa. 
The same applies to SLP Vectorization and loop vectorization by unrolling a loop.

RV aims to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
A region is defined as any control flow with a single entry and one or more non-divergent exits.
Non-divergent exits mean that control flow must not diverge to different exits starting from the 
same entry.
Such a region can then be, for example, an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.

To achieve this, RV implements partial control flow linearization~\cite{10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

@@

\section{Summary and Future Work}
\label{sec:summary}
In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model 
vectorization-specific instructions. 
This abstraction allows for efficiently exploring many optimization decisions simultaneously, 
keeping the underlying implementation flexible and modular. In the end, a cost model currently 
working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
and choose the most optimal one. Finally, this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
back into LLVM IR.
